---
title: "Homework 4 - Tidy data and joins"
author: "Hayden Scheiber -"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document
---

[Return to Main Page](https://github.com/HScheiber/STAT545-hw-Scheiber-Hayden/blob/master/README.md)

[Return to Homework 4 Landing Page](README.md)

---

- [General data reshaping and relationship to aggregation](#general-data-reshaping-and-relationship-to-aggregation)
    1. [Make a tibble with one row per year and columns for life expectancy for two or more countries](#make-a-tibble-with-one-row-per-year-and-columns-for-life-expectancy-for-two-or-more-countries)
    2. [Compute some measure of life expectancy for all possible combinations of continent and year](#compute-some-measure-of-life-expectancy-for-all-possible-combinations-of-continent-and-year)
    3. [Take a tibble with 24 rows: 2 per year, and reshape it so you have one row per year](#take-a-tibble-with-24-rows-2-per-year-and-reshape-it-so-you-have-one-row-per-year)

- [Join, merge, look up](#join-merge-look-up)
    1. [Create a second data frame, complementary to Gapminder. Join this with part of Gapminder](#create-a-second-data-frame-complementary-to-gapminder-join-this-with-part-of-gapminder)

---

Welcome! This is my attempt at developing my data wrangling and reshaping skills, as part of STAT 545 assignment 4.

First I need to load the `gapminder` dataset and the `tidyverse` package, as well as `knitr` for nicer table outputs. I also loaded `readr` to help import a comma separated data file.

```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(readr))
```

## General data reshaping and relationship to aggregation

### Make a tibble with one row per year and columns for life expectancy for two or more countries
<a href="#top">Back to top</a>

Let's compare life expectancy of __Canada__, __Rwanada__, and __China__. I begin by selecting data only from those countries using `filter`. Then, using `select` I retain only the columns I want to keep. I'll show the first 5 rows of this table to give an idea of what it looks like.

```{r}
data.selected.long <- gapminder %>%
        filter(country %in% c('Canada', 'Rwanda', 'China')) %>%
        select(country, year, lifeExp)

knitr::kable(head(data.selected.long, n = 5))
```

Now it's time to reshape this data into a different form. The `spread` function is used to convert a single column into multiple columns. This is useful to use on the __country__ column because there are only three values present there. The three values will become new columns in the reshaped dataframe.
```{r}
data.selected.wide <- data.selected.long %>%
        spread(country,lifeExp) 

knitr::kable(data.selected.wide,col.names = 
        c('Year','Life Expectancy in Canada (Years)', 
        'Life Expectancy in China (Years)', 
        'Life Expectancy in Rwanda (Years)'),
        align = 'c',
        caption = 'Based on data from Gapminder',
        format = 'markdown')
```

This is a very human-readable table. However, I found that it was easier to use the long format to make a scatterplot comparison of the selected country's life expectancies vs year.
```{r}
data.selected.long %>%
  ggplot(aes(x = year, y = lifeExp, colour=country)) + 
  geom_point() +
  scale_colour_discrete("Country") +
  scale_x_continuous(breaks = seq(1950, 2010, 10),
      labels = as.character(seq(1950, 2010, 10)),
      limits = c(1950, 2010),
      minor_breaks = NULL) +
  scale_y_continuous(breaks = seq(0, 85, 5),
      labels = as.character(seq(0, 85, 5)),
      limits = c(5, 85),
      minor_breaks = NULL) +
  theme_bw() + # black and white theme
  theme(axis.title = element_text(size=14),
      strip.text = element_text(size=14, face="bold"),
      plot.title = element_text(size=14, face="bold",hjust = 0.49),
      axis.text.x = element_text(size=12,face ="bold"),
      axis.text.y = element_text(size=12,face ="bold"),
      legend.title = element_text(size=14, face ="bold"),
      legend.text = element_text(size=12, face ="bold")) +
  labs(x = "Year", 
      y = "Life Expectancy",
      title = "Comparison of Life Expectancies: Canada, Rwanada, and China",
      caption = "Based on data from Gapminder")
```

### Compute some measure of life expectancy for all possible combinations of continent and year
<a href="#top">Back to top</a>

In this section, I decided to compute the median life expectancy for each continent at each year.
```{r,fig.width=8}
data.lifeExp.wide <- gapminder %>%
        group_by(continent, year) %>%
        mutate(lifeExp.median = median(lifeExp)) %>%
        select(continent, year, lifeExp.median) %>%
        unique() %>%
        spread(continent,lifeExp.median)

knitr::kable(data.lifeExp.wide,col.names = 
        c('Year','Africa', 
        'Americas', 
        'Asia', 
        'Europe', 
        'Oceania'),
        align = 'c',
        format = 'html', 
        caption = "<h4>Median Life Expectancy by Continent (Years)</h4>")

```
To exchange columns for rows is very easy in R. Just use the transpose function `t`.

```{r}
data.lifeExp.wide.t <- t(data.lifeExp.wide)
  
# This step forces the years to be formatted as integers instead of floating point numbers
data.lifeExp.wide.t.table <- rbind(formatC(data.lifeExp.wide.t[1,],format="d"),
        data.lifeExp.wide.t[-1,])

# Previous command removed the year label. Re-added it manually.
row.names(data.lifeExp.wide.t.table)[1] <- 'Year'

knitr::kable(data.lifeExp.wide.t.table, 
        col.names = NULL,
        align = 'c',
        format = 'html', 
        caption = "<h4>Median Life Expectancy by Continent (Years)</h4>")
```

This type of data format is human readable, but it isn't the best format for `ggplot`. The only types of plots I could _maybe_ see being easier with these formats are comparisons between different years or different continents. However, it seems to make colouring points by continent very awkward.
```{r,fig.width=8}
ggplot(data.lifeExp.wide,aes(x = year)) +
  geom_point(aes(y=Europe, colour='Europe')) +
  geom_point(aes(y=Asia, colour='Asia')) +
  scale_x_continuous(breaks = seq(1950, 2010, 10),
    labels = as.character(seq(1950, 2010, 10)),
    limits = c(1950, 2010),
    minor_breaks = NULL) +
  scale_colour_discrete(name = "Continent") +
  theme_minimal() +
  theme(axis.text.x = element_text(size=12,face ="bold"),
    axis.text.y = element_text(size=12,face ="bold"),
    plot.title = element_text(size=16,face ="bold",hjust = 0.5),
    axis.title = element_text(size=14, face ="bold"),
    legend.title = element_text(size=14, face ="bold"),
    legend.text = element_text(size=12, face ="bold")) +
  labs(x = "Year", 
    y = "Median Life Expectancy (Year)",
    title = "Comparison of Median Life Expectancy by Year: Asia vs Europe",
    caption = "Based on data from Gapminder")
```

If I use data in long format, as below, only a single layer of `geom_point` is required, but I also have to `filter` the data first. I think this method is better simply because of the colour aesthetic, which automatically maps to continent. In the previous example I had to manually map it, which just felt wrong.
```{r,fig.width=8}
# Same process as before, but without spread
data.lifeExp.long <- gapminder %>%
  group_by(continent, year) %>%
  mutate(lifeExp.median = median(lifeExp)) %>%
  select(continent, year, lifeExp.median) %>%
  unique()

data.lifeExp.long %>%
  filter(continent %in% c('Americas','Africa')) %>%
  ggplot(aes(x = year, y = lifeExp.median, colour = continent)) +
  geom_point() +
  scale_x_continuous(breaks = seq(1950, 2010, 10),
    labels = as.character(seq(1950, 2010, 10)),
    limits = c(1950, 2010),
    minor_breaks = NULL) +
  scale_colour_discrete(name = "Continent") +
  theme_minimal() +
  theme(axis.text.x = element_text(size=12,face ="bold"),
    axis.text.y = element_text(size=12,face ="bold"),
    plot.title = element_text(size=16,face ="bold",hjust = 0.5),
    axis.title = element_text(size=14, face ="bold"),
    legend.title = element_text(size=14, face ="bold"),
    legend.text = element_text(size=12, face ="bold")) +
  labs(x = "Year", 
    y = "Median Life Expectancy (Year)",
    title = "Comparison of Median Life Expectancy by Year: Americas vs Africa",
    caption = "Based on data from Gapminder")
```

### Take a tibble with 24 rows: 2 per year, and reshape it so you have one row per year
<a href="#top">Back to top</a>

For this, we start with the tibble created [here](http://stat545.com/block010_dplyr-end-single-table.html#window-functions "Window Functions").
```{r}
data.lifeExp.Asia <- gapminder %>%
  filter(continent == "Asia") %>%
  select(year, country, lifeExp) %>%
  group_by(year) %>%
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2) %>% 
  arrange(year)

knitr::kable(data.lifeExp.Asia, 
  col.names = c('Year','Country','Life Expectancy (Years)'),
  align = 'c',
  format = 'html', 
  caption = "<h4>Highest and Lowest Life Expectancy by Year in Asia</h4>")
```

I would like to reshape this data so that there is only one row per year. There must be an easier way of doing this, but the way I chose worked! First I added new columns that indicated the maximum and minimum life expectancy of each grouping of year, then cut away the previous columns with `select` and removed the now repeated data points with `unique`.
```{r}
data.lifeExp.Asia.grouped <- data.lifeExp.Asia %>%
  group_by(year) %>%
  mutate(min.lifeExp = min(lifeExp), 
         max.lifeExp = max(lifeExp),
         min.lifeExp.Country = country[1], 
         max.lifeExp.Country = country[2]) %>%
  select(year, 
         min.lifeExp.Country,
         min.lifeExp,
         max.lifeExp.Country,
         max.lifeExp) %>%
  unique()

knitr::kable(data.lifeExp.Asia.grouped,
  col.names = c("Year", 
                "Country with Lowest Life Expectancy",
                "Life Expectancy (Years)",
                "Country with Highest Life Expectancy",
                "Life Expectancy (Years)"),
  align = 'c',
  format = 'html', 
  caption = "<h4>Highest and Lowest Life Expectancy by Year in Asia (Years)</h4>")
```
This is what I wanted!

## Join, merge, look up


### Create a second data frame, complementary to Gapminder. Join this with part of Gapminder
<a href="#top">Back to top</a>

For this section, I downloaded a list of energy consumption per capita for 1960 to 2016 from the [world bank](https://data.worldbank.org/indicator/EG.USE.PCAP.KG.OE). I saved it as a `.csv` file and imported it with `read.csv`. Let's have a look at it's structure.
```{r}
data.energy <- read.csv("Energy_use.csv")

knitr::kable(head(data.energy), 
  align = 'c',
  format = 'html', 
  caption = "<h4>Energy Use in Kg of Oil Equivalent per Capita</h4>")
```

This is very ugly looking data! There are a few things wrong with it: I want to remove columns 2, 3, and 4. There's no data for 2016, so I'll remove that too. I also need to change the other headings so that they will match those in `gapminder`. After that's all done we can reshape it to match the shape of `gapminder`, and remove all those `NA` data points.
```{r}
# Remove columns 2:4 plus the last column for 2016
data.energy <- data.energy %>%
  select(ï..Country.Name, num_range("X", 1960:2015))

# Rename the columns
names(data.energy) <- c("country",1960:2015)

knitr::kable(head(data.energy), 
  align = 'c',
  format = 'html', 
  caption = "<h4>Energy Use in Kg of Oil Equivalent per Capita</h4>")
```

So now the data frame looks less ugly, but it still won't match up with `gapminder` as it is. We need to have the years consolidated into a single column, with countries repeated for each year. Looks like a job for `gather`. At this point I will also remove any missing data points, there's really no point in keeping them!
```{r}
data.energy.long <- data.energy %>%
  gather(year, Energy.Use.Per.Capita, 2:ncol(data.energy)) %>%
  filter(!is.na(Energy.Use.Per.Capita))

# Need to have year column as numeric
data.energy.long$year <- as.numeric(data.energy.long$year)

knitr::kable(head(data.energy.long,n=10), 
  align = 'c',
  format = 'html', 
  caption = "<h4>Energy Use in Kg of Oil Equivalent per Capita</h4>")
```

Finally, the data frame is in the shape I needed! Let's try joining it with gapminder using the various `join` functions.

#### Left join
I'll join this new data frame `data.energy.long` to gapminder using `left_join`. This type of join preserves all data in the first data frame (`gapminder` in this case) and adds new columns from the second data frame to matching rows. If there is no match for a row in the second data frame to the first, that data point will be excluded.
```{r}
gapminder.energy.left <- left_join(gapminder,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.left)
```

The warning comes about because the `country` columns in `gapminder` and `data.energy.long` are not just character vectors, but also have associated numbers: they are in the class `factor`. The numbers from the matching country strings do not necessarily match between these lists, resulting in a warning. We can convert the columns to character vectors with `as.character` to get rid of the warning.
```{r}
my.gap <- gapminder
my.gap$country <- as.character(my.gap$country)
data.energy.long$country <- as.character(data.energy.long$country)

gapminder.energy.left <- left_join(my.gap,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.left)
```

#### Right Join
Same as `left_join` but changed the arguments around.
```{r}
gapminder.energy.right <- right_join(my.gap,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.right)
```

So now all data from `data.energy.long` has been retained, and columns from gapminder have been added. Data from gapminder has been added only where it matches the country and year in `data.energy.long`.

#### Inner Join
`inner_join` retains only data found in both sets. There should be no empty data points here.
```{r}
gapminder.energy.inner <- inner_join(my.gap,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.inner)
```

#### Full Join
This function retains all data points from both data sets. This will result in the largest data frame, with lots of empty data points.
```{r}
gapminder.energy.full <- full_join(my.gap,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.full)
```

#### Semi Join
`semi_Join` is a way of filtering out the data in the first argument by keeping only the data points that have matches in the second argument. In this case, no columns from `Energy.user.Per.Capita` are added.
```{r}
gapminder.energy.semi <- semi_join(my.gap,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.semi)
```

#### Anti Join
This is another joining function. It is the inverse of `semi_join` and keeps only data points in the first argument _without_ matches in the second.
```{r}
gapminder.energy.anti <- anti_join(my.gap,data.energy.long, by = c("country","year"))

glimpse(gapminder.energy.anti)
```

That's all of the `join` functions!

 <a href="#top">Back to top</a>