---
title: "STAT 545 - Homework 2"
author: "Hayden Scheiber"
date: "September 21, 2017"
output: github_document
---

Welcome! This is an exploration of the Gapminder data frame, as part of STAT 545 assignment 2. Click [here](/hw02/README.md) to return to the HW02 readme, or [here](README.md) to return to the main repository.

First course of action is to load the dataset and the tidyverse package.

```{r}
library(gapminder)
library(tidyverse)
```

## Smell test the data: Exploring the Gapminder data set


#### Is it a data.frame, a matrix, a vector, a list?

Now, we want to know the structure of the data we're working with!

```{r}
str(gapminder)
```

It looks like gapminder is a dataframe, or "tibble" since we loaded tidyverse. Let's find out more:

```{r}
typeof(gapminder)
```
It seems that tibbles are a type of list.

#### What’s its class?
```{r}
class(gapminder)
```
So gapminder is a data frame, table, and tibble!

#### What data type is each variable?
```{r}
names(gapminder)
sapply(gapminder, class)
```

A healthy mix of various classes. And the data types:

```{r}
sapply(gapminder, typeof)
```


#### How many variables/columns? How many rows/observations?

Number of rows can be easily shown using the `nrow` function, whereas rows are shown with `ncol`

```{r}
nrow(gapminder)
ncol(gapminder)
```

#### Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different - contexts?

```{r}
dim(gapminder) #lists both dimensions of a data frame
length(gapminder) # also lists number of columns, may act differently for matrices
dim(unique(gapminder)) # shows number of unique entries
```

Which function you use will depend on the needs of the particular script! In particular, using the `unique()` function could be a very effective way of sorting through data that contains a lot of repeated features.

## Explore individual variables

Here we will take a closer look at a few of the variables in the gapminder dataset.

#### What are possible values (or range, whichever is appropriate) of each variable?

Lets begin with countries. How many are listed?

```{r}
unique(gapminder$country) %>%
  length()
```
So there are 142 countries in this dataset! We can list them nicely in alphabetical order with:

```{r}
unique(select(gapminder, country))
```

How about the range of years? What's in the earliest and latest year in the dataset?
```{r}
range(gapminder$year)
```
So the data set runs from 1952 to 2007.

Let's havea look at the life expectancy variable. How far does it range?

```{r}
range(gapminder$lifeExp)
```

That is quite the range! How about the range just in 2007, separately for each continent?

```{r}
filter(gapminder, year == 2007) %>%
  group_by(continent) %>%
    summarize(minLifeExp2007=min(lifeExp), maxLifeExp2007=max(lifeExp))
```


#### What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.

Let's take a look at a summary of the life expectancy:
```{r}
summary(gapminder$lifeExp)
```

We can plot max life expectancy vs year separately for each continent:
```{r}
group_by(gapminder,continent,year) %>%
  summarize(MaxLifeExp=max(lifeExp)) %>%
  ggplot(aes(x=year, y=MaxLifeExp)) + 
  geom_point(aes(colour=continent)) +
  labs(title = "Max Life Expectancy vs Year by Continent", x = "Year", 
       y = "Max Life Expectancy",caption = "Based on data from Gapminder")
```

Or this minimum life expectancy:
```{r}
group_by(gapminder,continent,year) %>%
  summarize(MinLifeExp=min(lifeExp)) %>%
  ggplot(aes(x=year, y=MinLifeExp)) + 
  geom_point(aes(colour=continent)) +
  labs(title = "Minimum Life Expectancy vs Year by Continent", x = "Year", 
       y = "Minimum Life Expectancy",caption = "Based on data from Gapminder")
```

## Explore various plot types

Why don't we try making a stripplot to see the spread of life expectancies for each continent. We can colour code by year, and overlay the average (red dot)!
```{r}
p <- ggplot(gapminder, aes(reorder(x = continent, lifeExp), y = lifeExp)) + 
  geom_jitter(aes(colour=year), position = position_jitter(width = 0.2, height = 0),alpha=0.2) +
  stat_summary(fun.y = mean, colour = "red", geom = "point", size = 3)
p + labs(title = "Life Expectancy In continents", x = "Continent", y = "Life Expectancy",caption = "Based on data from Gapminder")
```

That's pretty cool! A definite trend towards higher life expectancy as the years go on.

What about a histogram of life expectancy distribution in the 1990's, with colour overlay by continent?

```{r}
filter(gapminder, year > 1989 | year < 2000) %>%
  ggplot(aes(x = lifeExp,fill = continent)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Life Expectancy Distribution in the 90's", 
       x = "Life Expectancy", y = "Count",caption = "Based on data from Gapminder")
```

It might look better to plot this as a density plot:
```{r}
filter(gapminder, year > 1989 | year < 2000) %>%
  ggplot(aes(x = lifeExp,fill = continent)) +
  geom_density(alpha = 0.2) +
  labs(title = "Life Expectancy Distribution in the 90's", 
       x = "Life Expectancy", y = "Count",caption = "Based on data from Gapminder")
```

Okay, I think we've seen enough of life expectancy per year, how about life expectancy vs GDP per capita, in 2002, where size is determined by the population size?

```{r}
jPch <- 21
jDarkGray <- 'grey20'
jXlim <- c(150, 115000)
jYlim <- c(16, 100)

ggplot(subset(gapminder, year == 2002),
       aes(x = gdpPercap, y = lifeExp)) +
  scale_x_log10(limits = jXlim) + ylim(jYlim) +
  geom_point(aes(size = sqrt(pop)), pch = jPch, color = jDarkGray,
             show.legend = FALSE, alpha=0.5) + 
  scale_size_continuous(range=c(1,40)) +
  facet_wrap(~ continent) + coord_fixed(ratio = 1/43) +
  aes(fill = country) + scale_fill_manual(values = country_colors) +
  theme_bw() + theme(strip.text = element_text(size = rel(1.1))) + 
  labs(title = "Life Expectancy vs GDP per capita: 2002", 
      x = "GDP per Capita", y = "Life Expectancy",caption = "Based on data from Gapminder")
```

How about a bar graph of number of countries in each continent:
```{r}
group_by(gapminder,continent) %>%
  summarize(NumberofCountries = length(unique(country))) -> bardata

ggplot(bardata, aes(x = continent, weight=NumberofCountries)) + geom_bar(fill="Red") +
  labs(title = "Number of countries in the Continents", 
      x = "Continent", y = "Number of Countries",caption = "Based on data from Gapminder")

```

Lastly, here's a bar plot of total wealth in each continent in 2007:
```{r}
filter(gapminder, year == 2007) %>%
  group_by(country) %>%
  summarize(continent, CountryWealth = gdpPercap*pop) %>%
  group_by(continent) %>%
  summarize(TotalWealth = round( sum(CountryWealth)/1e12,digits = 1) ) -> bardata2

ggplot(bardata2, aes(x = continent, weight=TotalWealth)) + geom_bar(fill="Green") +
  labs(title = "Total Wealth in the Continents (2007)", 
      x = "Continent", y = "Total Wealth ($ Trillion)", caption = "Based on data from Gapminder")
```

## But I want to do more!

Here is some code. The analyst’s intent was to get the data for Rwanda and Afghanistan
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

This is not correct! Here are two possible ways to do it correctly:
```{r}
filter(gapminder, country %in% "Rwanda" | country == "Afghanistan")
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

