---
title: "Homework 3 - Gapminder Manipulation and Exploration with dplyr"
author: "Hayden Scheiber"
date: "September 26, 2017"
output: github_document
---

Welcome! This is an exploration of the Gapminder data frame, as part of STAT 545 assignment 2. Click [here](README.md) to return to the homework 3 landing page, or [here](https://github.com/HScheiber/STAT545-hw-Scheiber-Hayden/blob/master/README.md) to return repository home page.

First we need to load the `gapminder` dataset and the `tidyverse` package, as well as `knitr` for nicer table outputs.

```{r,warning=FALSE,message=FALSE,error=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
```

Here's an example of how to use `knitr::kable()` to make nicer looking tables. The `head()` command outputs the first `n` rows of an input data frame.
```{r results = 'asis'}
knitr::kable(head(gapminder,n=15))
```

## Task Menu

#### Get the maximum and minimum of GDP per capita for all continents.

This can be acheived easily using the `group_by()` combined with `summarize()`. Below, I group the gapminder data frame by continent and pipe that result into the summarize function, where I utilize the `max()` and `min()` functions to find the range of GDP per capita for each continent.
```{r}
p1 <- group_by(gapminder,continent) %>%
  summarize(Max_GDPperCap=max(gdpPercap), Min_GPDperCap = min(gdpPercap))
knitr::kable(p1)
```

Now we can plot this as a nice bar plot
```{r}
group_by(gapminder,continent) %>%
  summarize(NumberofCountries = length(unique(country))) -> bardata

ggplot(bardata, aes(x = continent, weight=NumberofCountries)) + geom_bar(fill="Red") +
  labs(title = "Number of countries in the Continents", 
      x = "Continent", y = "Number of Countries",caption = "Based on data from Gapminder")
```

